Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2016-01-16 WSSSPE3/report/wg_main_eng_design.tex
Update wg_main_eng_design.tex
Minor updates

This congregated list is an initial collection of principles that could be extended by adding from further related work form separate disciplines within the field of software engineering, including requirements engineering, software architecture, and testing. The group identified two example systems to discuss the application of the principles. The first one was a quantum chemistry system that allows the analysis of the characteristics and capabilities of molecules and solids. The second one was a modeling system for malaria that permitted biologists to analyze a range of datasets across geography, biology, and epidemiology, and add their own datasets. The group then examined the principles and took a retrospective analysis of what the developers did in practice against how the principles could have made a difference.

It was suggested that at the very minimum, the genetic building blocks of [technically] sustainable software should address two core quality attributes:
\begin{itemize}
\item Extensibility: the software’s ability to be extended and the level of effort required to implement the extension;
\item Maintainability: the effort required to locate and fix an error in operational software.
\end{itemize}
These fundamental building blocks could then be extended to include other quality attributes such as portability, resuability, scalability, usability, and energy efficiency etc. Nevertheless, this raises the question of what metrics and measures are suitable to demonstrate the sustainability of the software. This raised the question, What do the principles mean for scientific and computational software? Similalry, how do the principles relate to non-functional requirements? In addition, what do the five dimensions of sustainability mean for scientific software i.e. environmental, economic, social, techincal and individual?

>>>>

This congregated list is an initial collection of principles that could be extended by adding from further related work form separate disciplines within the field of software engineering, including requirements engineering, software architecture, and testing. The group identified two example systems to discuss the application of the principles. The first one was a quantum chemistry system that allows the analysis of the characteristics and capabilities of molecules and solids. The second one was a modeling system for malaria that permitted biologists to analyze a range of datasets across geography, biology, and epidemiology, and add their own datasets. The group then examined the principles and took a retrospective analysis of what the developers did in practice against how the principles could have made a difference. This raised the question, What do the principles mean for  computational scientific and engineering software? Similalry, how do the principles relate to non-functional requirements? It was suggested that at the very minimum, that sustainable software engineering principles should be mapped to two core quality attributes that underpin technically sustainable software:
\begin{itemize}
\item Extensibility: the software’s ability to be extended and the level of effort required to implement the extension;
\item Maintainability: the effort required to locate and fix an error in operational software.
\end{itemize}
These fundamental building blocks could then be extended to include other quality attributes such as portability, resuability, scalability, usability, and energy efficiency etc. Nevertheless, this raises the question of what metrics and measures are suitable to demonstrate the sustainability of the software. In addition, what do the five dimensions of sustainability mean for scientific software i.e. environmental, economic, social, techincal and individual?
